---
title: "Covid Veteran Data - Latent Profile Analysis"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r initialize, echo=FALSE, message=FALSE, warning=FALSE}
### Loading Libraries
# loading libraries to use functions for script

library(summarytools)
library(tidyverse)
library(reactable)
library(knitr)
library(car)
library(tidyLPA)

# show the code in the html output for all chunks 
# unless explicitly excepted.

knitr::opts_chunk$set(
   echo = TRUE,
   message = FALSE,
   warning = FALSE)
```

## Preprocessing

### Read VET longitudinal data
```{r}
rm(list=ls())

vetData <- read.csv("VET_timepoints_merged.csv", header = TRUE, na.strings=c("","NA"))
nrow(vetData)
```


### Create a unique id and assign to each participant
```{r}
vetData<-vetData%>%
  mutate(ID=1:n())%>%
  select(ID, everything())

#check for duplicates
vetData$ID[duplicated(vetData$ID)]
```

### Filter baseline data 
```{r}
#keep the baseline timepoint

vetData<-vetData %>% filter(redcap_event_name == "m0_arm_1")
```

### Mode function
```{r}
# define mode function
ModeFunc <- function(x, na.rm = FALSE) {
  if(na.rm){
    x = x[!is.na(x)]
  }
  ux <- unique(x)
  return(ux[which.max(tabulate(match(x, ux)))])
}
```

### Scale total calculation

### Moral Injury Outcome Scale (MIOS)

```{r MIOS}


#The total score range (14 items) is 0-56 for the total score and 0-28 for each of the two subscales (each of 7 items).


#### Shame subscale
colnmsShame = c("moi_mat_3", "moi_mat_8", "moi_mat_16", "moi_mat_18", "moi_mat_29", "moi_mat_32", "moi_mat_34")

#Calculate percent missing by dividing the number of NA cells by total cells, multiply by 100 and lastly round

vetData<- vetData %>% rowwise() %>%
  mutate(MIOSshame_perc_missing= round(((sum(is.na(c_across(all_of(colnmsShame))))/7)*100),2))

#calculate the mode of each row for the MIOS Shame items
vetData$MIOSshame_rowMode<- apply(vetData[,colnmsShame], 1, FUN = function(x) ModeFunc(na.omit(x)))

#impute missing values rowwise with rowmode for participants having less than 20% missing, then calculate scale total
vetData$MIOSshame_score<-NA
colnmsShame_plus<-c("ID", colnmsShame , "MIOSshame_perc_missing", "MIOSshame_rowMode")
MIOSshame_temp<-vetData[,colnmsShame_plus]

MIOSshame_temp <- MIOSshame_temp %>% filter(MIOSshame_perc_missing<=20) %>% rowwise() %>%
  mutate_at(vars(all_of(colnmsShame)), ~replace(., is.na(.), MIOSshame_rowMode))%>% 
  mutate(MIOSshame_score = mean(c_across(all_of(colnmsShame))))

vetData$MIOSshame_score[match(MIOSshame_temp$ID, vetData$ID)] <- MIOSshame_temp$MIOSshame_score
#test the results
# test_MIOSShame<- vetData %>% select(colnmsShame, MIOSshame_perc_missing, MIOSshame_rowMode, MIOSshame_score)


#### Trust-violation subscale

colnmsTrust = c("moi_mat_5", "moi_mat_6", "moi_mat_12", "moi_mat_14", "moi_mat_15", "moi_mat_20", "moi_mat_22")

#Calculate percent missing by dividing the number of NA cells by total cells, multiply by 100 and lastly round

vetData<- vetData %>% rowwise() %>%
  mutate(MIOStrust_perc_missing= round(((sum(is.na(c_across(all_of(colnmsTrust))))/7)*100),2))

#calculate the mode of each row for the MIOS Trust Violation-related items
vetData$MIOStrust_rowMode<- apply(vetData[,colnmsTrust], 1, FUN = function(x) ModeFunc(na.omit(x)))

#impute missing values rowwise with rowmode for participants having less than 20% missing, then calculate scale total
vetData$MIOStrust_score<-NA
colnmsTrust_plus<-c("ID", colnmsTrust , "MIOStrust_perc_missing", "MIOStrust_rowMode")
MIOStrust_temp<-vetData[,colnmsTrust_plus]


MIOStrust_temp <- MIOStrust_temp %>% filter(MIOStrust_perc_missing<=20) %>% rowwise() %>%
  mutate_at(vars(all_of(colnmsTrust)), ~replace(., is.na(.), MIOStrust_rowMode))%>% 
  mutate(MIOStrust_score = mean(c_across(all_of(colnmsTrust))))

vetData$MIOStrust_score[match(MIOStrust_temp$ID, vetData$ID)] <- MIOStrust_temp$MIOStrust_score

#test the results
# test_MIOStrust<- vetData %>% select(colnmsTrust, MIOStrust_perc_missing, MIOStrust_rowMode, MIOStrust_score)
#### MIOS Total Score

colnms_mios_14item = c(colnmsShame, colnmsTrust)

##mios_total_score
#Calculate percent missing by dividing the number of NA cells by total cells, multiply by 100 and lastly round.
vetData$mios_perc_missing <- round((rowSums(is.na(vetData[,colnms_mios_14item]))/14)*100, 2)

#calculate the mode of each row for the mios items (14 items)
vetData$mios_rowMode<- apply(vetData[,colnms_mios_14item], 1, FUN = function(x) ModeFunc(na.omit(x)))

#impute missing values rowwise with rowmode for participants having less than 20% missing, then calculate scale total

vetData$mios_score<-NA
colnms_mios_14item_plus<-c("ID", colnms_mios_14item , "mios_perc_missing", "mios_rowMode")
mios_temp<-vetData[,colnms_mios_14item_plus]

mios_temp <- mios_temp %>% filter(mios_perc_missing<=20) %>% rowwise() %>%
  mutate_at(vars(all_of(colnms_mios_14item)), ~replace(., is.na(.), mios_rowMode))%>% 
  mutate(mios_score = sum(c_across(all_of(colnms_mios_14item))))

vetData$mios_score[match(mios_temp$ID, vetData$ID)] <- mios_temp$mios_score

#test the results
# test_MIOS<- vetData %>% select(colnms_mios_14item, mios_rowMode, MIOStrust_rowMode, mios_score)

```

### Perform latent profile analysis on mios subscales
```{r}

mios_selection<-vetData %>%select(MIOSshame_score, MIOStrust_score)

#removing missing values
# df <- mios_selection[complete.cases(mios_selection), ]

mios_profile1<-mios_selection %>% estimate_profiles(1:5 , variances = "equal" ,covariances = "zero")
mios_profile1
plot_profiles(mios_profile1)
```

### Perform latent profile analysis on 14 mios items 
```{r}
#select all mios items

mios_temp<-mios_temp %>% select(all_of(colnms_mios_14item))

#removing missing values
# df <- mios_temp[complete.cases(mios_temp), ]

#perform latent profile analysis on 14 mios items
mios_profile2<-mios_temp %>% estimate_profiles(1:5 , variances = "equal" ,covariances = "zero")
mios_profile2
plot_profiles(mios_profile2)
```



### Appendix 

#### Entropy: A measure of classification uncertainty, reverse-coded so that 1 reflects complete certainty of classification, and 0 complete uncertainty.

#### BLRT p-value: p-value for the bootstrapped likelihood ratio test.
#### AIC: Aikake information criterion; based on -2 log-likelihood, and penalized by number of parameters.
#### BIC: Bayesian information criterion; based on -2 log-likelihood, and penalized by number of parameters adjusted by sample size.

#### Prob. Max.: Maximum of the diagonal of the average latent class probabilities for most likely class membership, by assigned class. The maximum should also be as high as possible, reflecting greater classification certainty (cases are assigned to classes they have a high probability of belonging to).
#### N Min.: Proportion of the sample assigned to the smallest class (based on most likely class membership).
#### N Max.: Proportion of the sample assigned to the largest class (based on most likely class membership).
